---
title: "Regression_Models_Course_Project"
author: "Craig Rowley, Nike, Inc."
date: "December 6, 2014"
output: pdf_document
---

Load model trends cars dataset and split into automatic transmission (am==0) and manual transmission (am==1) datasets.

```{r}
data(mtcars)
auto_trans <- mtcars[mtcars$am == 0,]
man_trans <- mtcars[mtcars$am == 1,]
```


Show that the MPG for manual transmission is greater than automatic transmission.

```{r}
mean(auto_trans$mpg)
mean(man_trans$mpg)
```


TODO: Need to break this apart into a physics problem.  Consider the Bias of the other influencing variables (e.g. weight, gears) and also the benefit of having more power (needs explored to see if this is true).  The goal is to find out if it is really the automatic transmission that is decreasing MPG.


#...EXample code from template
You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
