---
title: 'Reproducible Research, Week 3: Storm Data Research'
author: "Craig Rowley"
date: "December 17, 2014"
output:
  html_document:
    toc: true
    theme: united
---

#Synopsis

The following data analysis addresses the following questions for Extreme Weather Events in the United States:

- Which types of events are most harmful with respect to population health?
- Which types of events have the greatest economic consequences?

The analysis is based on data from the NOAA Storm Database, which contained more than 900K storm event records from 1950 through 2011 at the time of original research.  Details of the database are included in the following locations:

- National Weather Service [Storm Data Documentation](Storm Data Documentation)
- National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

For purposes of this analysis, injuries and deaths due to storms are used as the measure of harm to population health.  Total dollar impact to both property and crops are used to measure economic impact.

Our research shows that Tornados are responsible for nearly two-thirds of deaths caused by storms, while Floods account for almost one-third of all economic impact.  All steps to reach this conclusion are included below, including a detailed summarization in the Results section.


#Data Processing 

##Download file to ensure reproducibility:
```{r download, cache=TRUE}
bzip_filename <- "storm_data.csv.bz2"
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", bzip_filename, method="curl")

```

##Load file into memory for processing:
```{r load, cache=TRUE}
zz <- bzfile(bzip_filename, "r")
data <- read.csv(zz, header=TRUE)

```

##Create subset datasets specific to population health and economic consequence:
```{r subset, cache=TRUE}
to_dollar <- function(amount, code) { 
  #...R is by-value, so it is safe to cleanse the parameter directly
  code <- toupper(code)
  #...Convert Code to a multiplier and compute value
  #   NOTE: invalid/missing codes are assumed to be a multiplier of 1
  if (code=="K") {
    amount*1000
  } 
  else if (code=="M") {
    amount*1000*1000
  } 
  else if (code=="B") {
    amount*1000*1000*1000
  } 
  else {
    amount
  } 
}

health <- data.frame(
  event_type=data$EVTYPE, 
  fatalities=data$FATALITIES, 
  injuries=data$INJURIES, 
  total_health_impact=I(data$FATALITIES+data$INJURIES)
)

economic <- data.frame(
  event_type=data$EVTYPE, 
  property_dmg=data$PROPDMG,
  property_dmg_exp=data$PROPDMGEXP,
  crop_dmg=data$CROPDMG,
  crop_dmg_exp=data$CROPDMGEXP,
  total_economic_impact=apply( 
    data[,c('PROPDMG','PROPDMGEXP','CROPDMG','CROPDMGEXP')], 
    1, 
    function(x)
      to_dollar( as.numeric(x['PROPDMG']), x['PROPDMGEXP']) 
        +
      to_dollar( as.numeric(x['CROPDMG']), x['CROPDMGEXP']) 

    ) 
  )

```

##Aggregate the impact by event type and sort for Results output:
```{r aggregate, cache=TRUE}
#...Aggregate by store event type
total_health_impact_by_event_type <- tapply(health$total_health_impact, health$event_type, sum,  na.rm=TRUE)
mean_health_impact_by_event_type  <- tapply(health$total_health_impact, health$event_type, mean, na.rm=TRUE)

total_economic_impact_by_event_type <- tapply(economic$total_economic_impact, economic$event_type, sum,  na.rm=TRUE)
mean_economic_impact_by_event_type  <- tapply(economic$total_economic_impact, economic$event_type, mean, na.rm=TRUE)

#...Sort decreasing
total_health_impact_by_event_type   <- sort(total_health_impact_by_event_type,   decreasing=TRUE, na.last=TRUE)
mean_health_impact_by_event_type    <- sort(mean_health_impact_by_event_type,    decreasing=TRUE, na.last=TRUE)

total_economic_impact_by_event_type <- sort(total_economic_impact_by_event_type, decreasing=TRUE, na.last=TRUE)
mean_economic_impact_by_event_type  <- sort(mean_economic_impact_by_event_type,  decreasing=TRUE, na.last=TRUE)

#...Compute a percentage vector
pct <- function(x) {
  round( 100*x/sum(x), 2 )
}
pct_total_health_impact_by_event_type   <- pct(total_health_impact_by_event_type)
pct_mean_health_impact_by_event_type    <- pct(mean_health_impact_by_event_type)
pct_total_economic_impact_by_event_type <- pct(total_economic_impact_by_event_type)
pct_mean_economic_impact_by_event_type  <- pct(mean_economic_impact_by_event_type)


#...Glue this together for easy access in Results
ns <- function(x, els=10) {
  toupper( names(head(x,els)) )
}
vs <- function(x, els=10, precision=2) {
  as.vector(round(head(x,els),precision))
}
results <- data.frame(
  top_10_total_health_impact_name=ns(total_health_impact_by_event_type),
  top_10_total_health_impact_val=vs(total_health_impact_by_event_type),
  top_10_pct_total_health_impact_val=vs(pct_total_health_impact_by_event_type),

  top_10_mean_health_impact_name=ns(mean_health_impact_by_event_type),
  top_10_mean_health_impact_val=vs(mean_health_impact_by_event_type),
  top_10_pct_mean_health_impact_val=vs(pct_mean_health_impact_by_event_type),

  top_10_total_economic_impact_name=ns(total_economic_impact_by_event_type),
  top_10_total_economic_impact_val=vs(total_economic_impact_by_event_type),
  top_10_pct_total_economic_impact_val=vs(pct_total_economic_impact_by_event_type),

  top_10_mean_economic_impact_name=ns(mean_economic_impact_by_event_type),
  top_10_mean_economic_impact_val=vs(mean_economic_impact_by_event_type),
  top_10_pct_mean_economic_impact_val=vs(pct_mean_economic_impact_by_event_type)

  )

```

##Create useful variables for the Results section
```{r variables, cache=TRUE}
dates <- data$BGN_DATE[!is.na(data$BGN_DATE)]
mindate <- format( min(as.Date(dates,"%m/%d/%Y 0:00:00")), format="%B %d, %Y" )
maxdate <- format( max(as.Date(dates,"%m/%d/%Y 0:00:00")), format="%B %d, %Y" )
stormEventCount <- nrow(data)
stormEventTypeCount <- length( levels(data$EVTYPE))

yearCount = length( levels(as.factor( format( as.Date(dates,"%m/%d/%Y 0:00:00"), format="%Y" )) ))

totalHealthImpact <- format(sum(total_health_impact_by_event_type), digits=2, big.mark=",")

mostDeadlyStorm <- data[data$FATALITIES==max(data$FATALITIES),]
mostDeadlyStormDate <- format( as.Date(mostDeadlyStorm$BGN_DATE, "%m/%d/%Y 0:00:00"), format="%B %d, %Y" )

numTotalEconomicImpact <- sum(total_economic_impact_by_event_type)/(1000*1000*1000)
totalEconomicImpact <- format( numTotalEconomicImpact, digits=2, big.mark="," )
avgDollarPerYear <- format( numTotalEconomicImpact/yearCount, digits=2, big.mark=",")

top3pct <- sum(results$top_10_pct_total_economic_impact_val[1:3])
top3pctEcon <- format( top3pct, digits=2, big.mark=",")

```

#Results

The NOAA storm database contains `r format(stormEventCount, digits=2, big.mark=",")` storm event records for `r stormEventTypeCount` types of storms from `r mindate` to `r maxdate`.


##Health Impact Summary

This research states that Tornados cause more deaths than all other storm types combined (Figure 1).  However, Heat Waves, on average, are the most deadly events (Figure 2).

In total, `r totalHealthImpact` people were injured or killed by storms across `r yearCount` years.  The most deadly storm occurred in `r mostDeadlyStorm$BGN_LOCATI` on `r mostDeadlyStormDate` and killed `r mostDeadlyStorm$FATALITIES` people.  Perhaps surprisingly, this most deadly event was a Heat Wave.


###Figure 1: Top 10 Weather Event Types by Total Health Impact (Persons)
```{r output_health_total, cache=TRUE}
library(knitr)

d <-  data.frame(
  "Type"=results$top_10_total_health_impact_name, 
  "Total Impact"=results$top_10_total_health_impact_val, 
  "Percentage"=results$top_10_pct_total_health_impact_val
  )
kable(d, format="markdown")
```

###Figure 2: Top 10 Weather Event Types by Mean Average Health Impact (Persons)
```{r output_health_mean, cache=TRUE}
d <-  data.frame(
  "Type"=results$top_10_mean_health_impact_name, 
  "Average Impact"=results$top_10_mean_health_impact_val, 
  "Percentage"=results$top_10_pct_mean_health_impact_val
  )
kable(d, format="markdown")


```


##Economic Impact

Floods may not kill as many people, but they have more than double the economic impact of any other storm type.

Storms caused more than `r totalEconomicImpact` Billion dollars of economic impact across `r yearCount` years.  That is more than `r avgDollarPerYear` Billion dollars on average, per year (which exceeds the [GDP of Rwanda in 2013](http://en.wikipedia.org/wiki/List_of_countries_by_GDP_%28nominal%29)).

Figure 2 shows the top 10 weather event types, along with their total economic impact and the overall percentage impact.  Floods, Hurricanes, Typhoons, and Tornados account for more than `r top3pctEcon`%  of overall economic impact.


###Figure 3: Top 10 Weather Event Types by Total Economic Impact (Billions)
```{r output_economic, cache=TRUE}
d <-  data.frame(
  "Type"=results$top_10_total_economic_impact_name, 
  "Impact"=round(results$top_10_total_economic_impact_val/(1000*1000*1000),2), 
  "Percentage"=results$top_10_pct_total_economic_impact_val
  )
kable(d, format="markdown")

```
